---
title: "Impact of Sever Weather on Public Health and Econmics"
author: "Craig Tedeton"
date: "06/04/2016"
output: github_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Synopsis

The goal of this research is to explore the NOAA Storm Database and potenetial impact of 
severe weather events on both the population and economy.

The database covers the time period from between 1950 and November 2012. In the earlier years 
of the database there are generally fewer events recorded, most likely due to a lack of good records. 
More recent years should be considered more complete.

The analysis aims to investigate which different types of sever weather events are most harmful on the 
populations health in respect of general injuries and fatalities. Further the economic consequences will 
be analyzed by exploring the financial damage done to both general property and agriculture (i.e. crops)

2. Data Processing
The data being processed was gathered from the NOAA Storm Database. It was accessed and processed utilizing the following code:

```{r echo=TRUE}
# Data Processing
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, "./data/StormData.csv.bz2")
fpath <- file.path("./data")
filelist <- list.files(fpath, recursive = TRUE)
filelist
fileName <- "./data/StormData.csv.bz2"

stormData <- read.csv(fileName)
summary(stormData)
rows <- nrow(stormData) ; cols <- ncol(stormData)
rows ; cols
```

2.1 Subset of Data
We will be looking specifically at a subset of this data. So, we need to extract 
the columns for property damage, injuries and fatalities
```{r echo=TRUE}
eventData <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
myData <- stormData[eventData]

```
2.1.1 Finding Property Damage 
The data also requires we find unique information on property damage, thus we will not look
at the injuries and fatalities as of yet. Invalid data was excluded by assigning a value of zero (‘0’). 
Then property damage value was calculated by multiplying the property damage and property exponent.
```{r echo=TRUE}
unique(myData$PROPDMGEXP)

# assign value to property exponent
myData$PROPEXP[myData$PROPDMGEXP == "K"] <- 1000
myData$PROPEXP[myData$PROPDMGEXP == "M"] <- 1e+06
myData$PROPEXP[myData$PROPDMGEXP == ""] <- 1
myData$PROPEXP[myData$PROPDMGEXP == "B"] <- 1e+09
myData$PROPEXP[myData$PROPDMGEXP == "m"] <- 1e+06
myData$PROPEXP[myData$PROPDMGEXP == "0"] <- 1
myData$PROPEXP[myData$PROPDMGEXP == "5"] <- 1e+05
myData$PROPEXP[myData$PROPDMGEXP == "6"] <- 1e+06
myData$PROPEXP[myData$PROPDMGEXP == "4"] <- 10000
myData$PROPEXP[myData$PROPDMGEXP == "2"] <- 100
myData$PROPEXP[myData$PROPDMGEXP == "3"] <- 1000
myData$PROPEXP[myData$PROPDMGEXP == "h"] <- 100
myData$PROPEXP[myData$PROPDMGEXP == "7"] <- 1e+07
myData$PROPEXP[myData$PROPDMGEXP == "H"] <- 100
myData$PROPEXP[myData$PROPDMGEXP == "1"] <- 10
myData$PROPEXP[myData$PROPDMGEXP == "8"] <- 1e+08

# handling invalid data
myData$PROPEXP[myData$PROPDMGEXP == "+"] <- 0
myData$PROPEXP[myData$PROPDMGEXP == "-"] <- 0
myData$PROPEXP[myData$PROPDMGEXP == "?"] <- 0

# calculate property damage
myData$PROPDMGVAL <- myData$PROPDMG * myData$PROPEXP
```

2.1.2 Finding Crop Damages
Crop damage exponent by level were listed and assigned for the crop exponent data. 
Invalid data was excluded by assigning the value of zero (‘0’). Then crop damage value was calculated by 
multiplying the crop damage and crop exponent.
```{r echo=FALSE}
# exploring the crop exponent data
unique(myData$CROPDMGEXP)

# assign values to crop xponents
myData$CROPEXP[myData$CROPDMGEXP == "M"] <- 1e+06
myData$CROPEXP[myData$CROPDMGEXP == "K"] <- 1000
myData$CROPEXP[myData$CROPDMGEXP == "m"] <- 1e+06
myData$CROPEXP[myData$CROPDMGEXP == "B"] <- 1e+09
myData$CROPEXP[myData$CROPDMGEXP == "0"] <- 1
myData$CROPEXP[myData$CROPDMGEXP == "k"] <- 1000
myData$CROPEXP[myData$CROPDMGEXP == "2"] <- 100
myData$CROPEXP[myData$CROPDMGEXP == ""] <- 1

# handling invalid data
myData$CROPEXP[myData$CROPDMGEXP == "?"] <- 0

# calculate crop damage
myData$CROPDMGVAL <- myData$CROPDMG * myData$CROPEXP
```

2.2 Calculating the Costs of the Events

Observations note that the " most harmful to population health" events are fatalities and injuries. So those are the only ones selected.

Observations note that the " most harmful to economic problem“” events are property and crop damage. So those events are the only selected.

For each incident (fatalities, Injuries, property and crop damage), the total values were estimated.

We also need to aggregate to values for the events

```{r echo=TRUE}
# totals by events
fatality <- aggregate(FATALITIES ~ EVTYPE, myData, FUN = sum)
injury <- aggregate(INJURIES ~ EVTYPE, myData, FUN = sum)
propDamag <- aggregate(PROPDMGVAL ~ EVTYPE, myData, FUN = sum)
cropDamag <- aggregate(CROPDMGVAL ~ EVTYPE, myData, FUN = sum)
```

2.3 Reviewing Highest Fatalities and Injuries

2.3.1 Top 10 Highest Fatalities and Injuries

Plotting the top 10 highest fatalities and injuries will help better your understanding of the impact of these observations.

```{r echo=TRUE}
# highest fatalities
fatalsTop10 <- fatality[order(-fatality$FATALITIES), ][1:10, ]

# highest injuries
injuryTop10 <- injury[order(-injury$INJURIES), ][1:10, ]

# plots of Top 10s
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)

barplot(fatalsTop10$FATALITIES, las = 3, names.arg = fatalsTop10$EVTYPE, main = "Top 10 Events-Highest Fatalities", 
        ylab = "Number of Fatalities", col = "red")

barplot(injuryTop10$INJURIES, las = 3, names.arg = injuryTop10$EVTYPE, main = "Top 10 Events-Highest Injuries", 
        ylab = "Number of Injuries", col = "orange")

```

2.3.2 Top 10 Highest Property and Crop Damages

Plotting the top 10 highest fatalities and injuries will help better your understanding of the impact of these observations.


```{r echo=TRUE}
# highest property damage
propDamag10 <- propDamag[order(-propDamag$PROPDMGVAL), ][1:10, ]

# highest crop damage
cropDamag10 <- cropDamag[order(-cropDamag$CROPDMGVAL), ][1:10, ]

par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)

barplot(propDamag10$PROPDMGVAL/(10^9), las = 3, names.arg = propDamag10$EVTYPE, main = "Events with Highest Property Damage", 
        ylab = "Costs ($ Bil)", col = "brown")

barplot(cropDamag10$CROPDMGVAL/(10^9), las = 3, names.arg = cropDamag10$EVTYPE, main = "Events With Highest Crop Damage", 
        ylab = "Costs ($ Bil)", col = "darkgreen")
```
<br>
3. Results of the Research

After review of the data observations and the plotting of this information, we may conclude that;
Their is a correlation with the highest number of fatalities and injuries with the TORNADO event. Also, excessive heat was noteworthy for high fatalities, while thunderstorm wind appears to contribute to injuries.

Their is also an apparent correlation between floods and maximum property damage. While drought caused maximum crop damage with flood also being highly noted.

